Data 621---HW1---Regression Model---Version 2

### 2020-09-09 HW1/RegModel_Adler.Rmd
Copyedit and add R2/aR2 to test set and table to training

# Meltng data from wide to narrow for munging & ggplot
MBmelt <- melt(MB, id.vars = 'INDEX', variable.name = 'Metric',
               value.name = 'Value', variable.factor = FALSE)
```

@@

# Use the Freedman-Diaconis rule for historgram bin widths instead of ggplot
# defaults

@@

### Training and Test Sets
As we are performing this analysis manually, we will break up the supplied data
into two sets, one for training and one for testing. This needs to be done
before any normalisation or imputation, otherwise the testing set is polluted
with information from the training set. Please note that obervations prior to
this point were on the entirety of the data set.

```{r testTrain}
set.seed(811247)
trainObs <- sample(x = n, size = 0.75 * n, replace = FALSE)
MBtrn <- MB[trainObs]
MBtst <- MB[!trainObs]
MBtrnlng <- melt(MBtrn, id.vars = 'INDEX', variable.name = 'Metric',
                 value.name = 'Value', variable.factor = FALSE)
MBtstlng <- melt(MBtst, id.vars = 'INDEX', variable.name = 'Metric',
                 value.name = 'Value', variable.factor = FALSE)

@@

## Analysis
Normally, we would use a statistical package like `caret`, which not only
simplifies model construction but also takes of splitting into training and
test sets and takes care of cross-validation. For the purposes of this homework,
and to demonstrate understanding of the underlying decisions involved in
predictive modeling, we will use calls to base R and the `MASS` package.

As the transforms were imposed on the entirety of the training data, we ar

@@

 1. The null model: This is the model with only the intercept, the grand mean.
 2. The saturated model: this is the model with every term included
 3. A model allowing pairwise interactions within the data types of baserunning,
 batting, pitching, and fielding, fit using a forward and backwards stepwise
 regression, and the model
resulting from the stepwise regression. A nuFor the regression, we will consider at
most pairwise interactions within the classes of metrics.

@@

# Rscale TARGET_WINS to its standard value.
MBtrnSC[, TARGET_WINS := TARGET_WINS * 
          MBtrnlngUnscale[Metric == 'TARGET_WINS', SD] +
          MBtrnlngUnscale[Metric == 'TARGET_WINS', Mean]]

@@

### Discussion
The first item for discussion is the observation that while the intercept does
not change between the null model and the saturated model, it does decrease for
the model allowing within class pairwise interactions. There are 162 games in a
baseball season. According to all the models, given no other information, a team
will lose slightly more than half its games. Apparently, interctions in the
stepwise model unearth somewhat more information about why teams win.

@@

#### Residual Plots
```{r residPlots, fig.width=7}
par(mfrow = c(2, 2))
plot(saturatedModel$fitted.values, studres(saturatedModel),
     main = 'Saturated Model', xlab = 'Predicted Values',
     ylab = 'Studentized Residuals', xlim = c(0, 160), ylim = c(-6, 6))
plot(stepModel$fitted.values, studres(stepModel),
     main = 'Stepwise Pairwise Model', xlab = 'Predicted Values',
     ylab = 'Studentized Residuals', xlim = c(0, 160), ylim = c(-6, 6))
plot(nullModel$fitted.values, studres(nullModel),
     main = 'Null Model', xlab = 'Predicted Values',
     ylab = 'Studentized Residuals', xlim = c(0, 160), ylim = c(-6, 6))
par(mfrow = c(1, 1))
```

Both the saturated and the stepwise models look to have a basically normal
distribution of their residuals. The null model returns a single value,
explaining the bizarness of the graph.

The **actual** model to be used for the prediction will be the model which
returns the best RMSE and MAPE on the test set.

@@

 * Reasonable
   * Caught stealing removes baserunners; negative coefficient makes sense.
   * Stolen bases get a runner closer to home plate; positive coefficient makes
   sense.
   * Triples get a runner very close to home plate; positive coefficient makes
   sense.
   * Walks are free baserunners; positive coefficient makes sense.
   * Hits increase the number of baserunners; positive coefficient makes sense.
   * Hitting home runs directly increase the score; positive coefficient makes
   sense.
   * Striking out reduces the number of baserunners; negative coefficient makes
   sense.
   * Errors allow the other team free baserunners; negative coefficient makes
   sense.
   * Giving up home runs gives the opponent scores; negative coefficient makes
   sense.
   * Getting strikeouts reduces the opponents baserunners; positive coefficient
   makes sense.
 * Curious
   * Hitting doubles get runners on base; why negative?
   * Turning double plays reduce baserunners; why negative?
   * Giving up walks allows the other team free baserunners; negative
   coefficient makes sense but why insignificant?
   * Giving up hits allows the other team baserunners; why positive athough
   insignificant?

An absolutely fascinating observation. In the first draft of this exercise,
`TEAM_BATTING_HR` was removed due to its high correlation with
`TEAM_PITCHING_HR`. Note the coefficients, batting is almost +12 and pitching is
-9. However, this is such as strong indicator, that in the first run, pitching
was given a *postive* coefficient of around +3. In hindsight this is because
*it was being used as an indicator for* ***batting!!*** The correlation allowed
the use of pitching HRs as an indicator for the hidden batting HRs! Once both
were restored to the model, the logical coefficients surfaced. Another reason
why models should not be trusted out of the box, but all model results should
be reviewed for sanity and sense!!

@@

A possible explanation for the negative coefficent for getting double plays, is
that double plays require at least two people on base. That means that the
opponent has a lot of base runners, which is very highly correlated with
scoring.

@@

General inferences about the nature of the coefficients is in the previous
section. The fascinating example of how multi-collinearity can override logic is
also discussed in the previous secton.

@@

# Center and Scale
MBtstlngSC[, Value := replace(Value, is.na(Value), mean(Value, na.rm = TRUE)),
       keyby = Metric]
MBtstlngUnscale <- MBtstlngSC[, .(Mean = mean(Value), SD = sd(Value)),
                          keyby = Metric]
MBtstlngSC[, Value := scale(Value, center = TRUE, scale = TRUE), keyby = Metric]
MBtstSC <- dcast(MBtstlngSC, INDEX ~ Metric, value.var = 'Value')
MBtstSC[, TARGET_WINS := TARGET_WINS * 
          MBtstlngUnscale[Metric == 'TARGET_WINS', SD] +
          MBtstlngUnscale[Metric == 'TARGET_WINS', Mean]]
nmP <- predict(nullModel, MBtstSC)
smP <- predict(saturatedModel, MBtstSC)
stmP <- predict(stepModel, MBtstSC)
nmRMSE <- sqrt(mean((MBtstSC$TARGET_WINS - nmP) ^ 2))
nmMAPE <- mean(abs((MBtstSC$TARGET_WINS - nmP) / MBtstSC$TARGET_WINS))
smRMSE <- sqrt(mean((MBtstSC$TARGET_WINS - smP) ^ 2))
smMAPE <- mean(abs((MBtstSC$TARGET_WINS - smP) / MBtstSC$TARGET_WINS))
stmRMSE <- sqrt(mean((MBtstSC$TARGET_WINS - stmP) ^ 2))
stmMAPE <- mean(abs((MBtstSC$TARGET_WINS - stmP) / MBtstSC$TARGET_WINS))
knitr::kable(data.frame(Model = c("Null", "Saturated", "Stepwise Pairwise"),
                        RMSE = c(nmRMSE, smRMSE, stmRMSE),
                        MAPE = c(nmMAPE, smMAPE, stmMAPE)),
             digits = 3L, caption = 'Test Set Results')
```

The stepwise model allowing for pairwise interactions is the winner in both
cases, as was expected.

### Predictions
We will use the stepwise-pairwise model to predict `TARGET_WINS` for the
separately provided evaluation set by following the same preprocessing used for
the training and testing sets.

@@

# Appendix
The code chunks below represent the code called in order during the analysis.
They are reproduced in the appendix for review and comment.

```{r loadPackages, echo=TRUE, eval=FALSE}
```
```{r readData, echo=TRUE, eval=FALSE}
```
```{r missingTable, echo=TRUE, eval=FALSE}
```
```{r summaryTable, echo=TRUE, eval=FALSE}
```
```{r boxPlot1, echo=TRUE, eval=FALSE}
```
```{r boxPlot2, echo=TRUE, eval=FALSE}
```
```{r corrgram, echo=TRUE, eval=FALSE}
```
```{r testTrain, echo=TRUE, eval=FALSE}
```
```{r imputeMean, echo=TRUE, eval=FALSE}
```
```{r centerScale, echo=TRUE, eval=FALSE}
```
```{r plotScaled, echo=TRUE, eval=FALSE}
```
```{r boxCox, echo=TRUE, eval=FALSE}
```
```{r reScale, echo=TRUE, eval=FALSE}
```
```{r reImputeThenScale, echo=TRUE, eval=FALSE}
```
```{r lm, echo=TRUE, eval=FALSE}
```
```{r nullOutput, echo=TRUE, eval=FALSE}
```
```{r satOutput, echo=TRUE, eval=FALSE}
```
```{r stepOutput, echo=TRUE, eval=FALSE}
```
```{r residPlots, echo=TRUE, eval=FALSE}
```
```{r buildTestSet, echo=TRUE, eval=FALSE}
```
```{r predictions, echo=TRUE, eval=FALSE}
```

>>>>

# Melting data from wide to narrow for munging & ggplot
MBmelt <- melt(MB, id.vars = 'INDEX', variable.name = 'Metric',
               value.name = 'Value', variable.factor = FALSE)
```

@@

# Use the Freedman-Diaconis rule for histogram bin widths instead of ggplot
# defaults

@@

### Training and Test Sets
As we are performing this analysis manually, we will break up the supplied data
into two sets, one for training and one for testing. This needs to be done
before any normalization or imputation, otherwise the testing set is polluted
with information from the training set. Please note that observations prior to
this point were on the entirety of the data set.

```{r testTrain}
set.seed(811247)
trainObs <- sample(x = n, size = 0.75 * n, replace = FALSE)
MBtrn <- MB[trainObs]
MBtst <- MB[!trainObs]
ntrn <- dim(MBtrn)[[1]]
ntst <- dim(MBtst)[[1]]
MBtrnlng <- melt(MBtrn, id.vars = 'INDEX', variable.name = 'Metric',
                 value.name = 'Value', variable.factor = FALSE)
MBtstlng <- melt(MBtst, id.vars = 'INDEX', variable.name = 'Metric',
                 value.name = 'Value', variable.factor = FALSE)

@@

## Analysis
Normally, we would use a statistical package like `caret`, which not only
simplifies model construction but also takes care of cross-validation. For the
purposes of this homework, and to demonstrate understanding of the underlying
decisions involved in predictive modeling, we will use calls to base R and the
`MASS` package.

@@

 1. The null model: This is the model with only the intercept, the grand mean.
 2. The saturated model: this is the model with every term included.
 3. A model allowing pairwise interactions within the data types of baserunning,
 batting, pitching, and fielding, fit using a forward and backwards stepwise
 regression.

@@

# Rescale TARGET_WINS to its standard value.
MBtrnSC[, TARGET_WINS := TARGET_WINS * 
          MBtrnlngUnscale[Metric == 'TARGET_WINS', SD] +
          MBtrnlngUnscale[Metric == 'TARGET_WINS', Mean]]

@@

### Discussion
The first item for discussion is the observation that while the intercept does
not change between the null model and the saturated model, it does decrease for
the model allowing within class pairwise interactions. There are 162 games in a
baseball season. According to all the models, given no other information, a team
will lose slightly more than half its games. Apparently, interactions in the
stepwise model unearth somewhat more information about why teams win.

@@

```{r trainStatTable}
knitr::kable(data.frame(Model = c("Null", "Saturated", "Stepwise Pairwise"),
                        Rsq = c(snm$r.squared, ssm$r.squared, sstm$r.squared),
                        AdjRsq = c(snm$adj.r.squared, ssm$adj.r.squared,
                                   ssm$adj.r.squared),
                        RMSE = c(sqrt(mean(nullModel$residuals ^ 2)),
                                 sqrt(mean(saturatedModel$residuals ^ 2)),
                                 sqrt(mean(stepModel$residuals ^ 2)))),
             digits = 3L, caption = 'Train Set Results')
```

#### Residual Plots
```{r residPlots, fig.height=7}
par(mfrow = c(2, 2))
plot(saturatedModel$fitted.values, studres(saturatedModel),
     main = 'Saturated Model', xlab = 'Predicted Values',
     ylab = 'Studentized Residuals', xlim = c(0, 160), ylim = c(-6, 6))
plot(stepModel$fitted.values, studres(stepModel),
     main = 'Stepwise Pairwise Model', xlab = 'Predicted Values',
     ylab = 'Studentized Residuals', xlim = c(0, 160), ylim = c(-6, 6))
plot(nullModel$fitted.values, studres(nullModel),
     main = 'Null Model', xlab = 'Predicted Values',
     ylab = 'Studentized Residuals', xlim = c(0, 160), ylim = c(-6, 6))
par(mfrow = c(1, 1))
```

Both the saturated and the stepwise models look to have a basically normal
distribution of their residuals. The null model returns a single value,
explaining the bizarreness of the graph.

The **actual** model to be used for the prediction will be the model which
returns the best RMSE on the test set. Unfortunately, MAPE cannot be used as
there are observations which are 0.

@@

 * Reasonable
   * Caught stealing removes baserunners; negative coefficient makes sense.
   * Stolen bases get a runner closer to home plate; positive coefficient makes
   sense.
   * Triples get a runner very close to home plate; positive coefficient makes
   sense.
   * Walks are free baserunners; positive coefficient makes sense.
   * Hits increase the number of baserunners; positive coefficient makes sense.
   * Hitting home runs directly increase the score; positive coefficient makes
   sense.
   * Striking out reduces the number of baserunners; negative coefficient makes
   sense.
   * Errors allow the other team free baserunners; negative coefficient makes
   sense.
   * Giving up home runs gives the opponent scores; negative coefficient makes
   sense.
   * Getting strikeouts reduces the opponents baserunners; positive coefficient
   makes sense.
 * Curious
   * Hitting doubles get runners on base; why negative?
   * Turning double plays reduce baserunners; why negative?
   * Giving up walks allows the other team free baserunners; negative
   coefficient makes sense but why insignificant?
   * Giving up hits allows the other team baserunners; why positive although
   insignificant?

An absolutely fascinating observation. In the first draft of this exercise,
`TEAM_BATTING_HR` was removed due to its high correlation with
`TEAM_PITCHING_HR`. Note the coefficients, batting is almost +12 and pitching is
-9. However, this is such as strong indicator, that in the first run, pitching
was given a *positive* coefficient of around +3. In hindsight this is because
*it was being used as an indicator for* ***batting!!*** The correlation allowed
the use of pitching HRs as an indicator for the hidden batting HRs! Once both
were restored to the model, the logical coefficients surfaced. Another reason
why models should not be trusted out of the box, but all model results should
be reviewed for sanity and sense!!

@@

A possible explanation for the negative coefficient for getting double plays, is
that double plays require at least two people on base. That means that the
opponent has a lot of base runners, which is very highly correlated with
scoring.

@@

General inferences about the nature of the coefficients is in the previous
section. The fascinating example of how multi-collinearity can override logic is
also discussed in the previous section.

@@

# Center and Scale
MBtstlngSC[, Value := replace(Value, is.na(Value), mean(Value, na.rm = TRUE)),
       keyby = Metric]
MBtstlngUnscale <- MBtstlngSC[, .(Mean = mean(Value), SD = sd(Value)),
                          keyby = Metric]
MBtstlngSC[, Value := scale(Value, center = TRUE, scale = TRUE), keyby = Metric]
MBtstSC <- dcast(MBtstlngSC, INDEX ~ Metric, value.var = 'Value')
MBtstSC[, TARGET_WINS := TARGET_WINS * 
          MBtstlngUnscale[Metric == 'TARGET_WINS', SD] +
          MBtstlngUnscale[Metric == 'TARGET_WINS', Mean]]
nmP <- predict(nullModel, MBtstSC)
nmPR <- (MBtstSC$TARGET_WINS - nmP)
smP <- predict(saturatedModel, MBtstSC)
smPR <- (MBtstSC$TARGET_WINS - smP)
stmP <- predict(stepModel, MBtstSC)
stmPR <- (MBtstSC$TARGET_WINS - stmP)
SST <- sum((MBtstSC$TARGET_WINS - mean(MBtstSC$TARGET_WINS)) ^ 2)
nmPR2 <- 1 - sum(nmPR ^ 2) / SST
nmPaR2 <- 1 - (1 - nmPR2) * (snm$df[[1]] + snm$df[[2]] - 1) / (snm$df[[2]])
nmPRMSE <- sqrt(mean(nmPR ^ 2))

smPR2 <- 1 - sum(smPR ^ 2) / SST
smPaR2 <- 1 - (1 - smPR2) * (ssm$df[[1]] + ssm$df[[2]] - 1) / (ssm$df[[2]])
smPRMSE <- sqrt(mean(smPR ^ 2))

stmPR2 <- 1 - sum(stmPR ^ 2) / SST
stmPaR2 <- 1 - (1 - stmPR2) * (sstm$df[[1]] + sstm$df[[2]] - 1) / (sstm$df[[2]])
stmPRMSE <- sqrt(mean(stmPR ^ 2))
knitr::kable(data.frame(Model = c("Null", "Saturated", "Stepwise Pairwise"),
                        Rsq = c(nmPR2, smPR2, stmPR2),
                        AdjRsq = c(nmPaR2, smPaR2, stmPaR2),
                        RMSE = c(nmPRMSE, smPRMSE, stmPRMSE)),
             digits = 3L, caption = 'Test Set Results')
```

The stepwise model allowing for pairwise interactions is the winner in all
cases, as was expected. The metrics are all worse than those on the training
set. This is a classic example of overfitting, which may have been helped by
k-fold cross-validation.

Also, this is one of those rare times where the actual \(R^2\) can be negative.
The null model prediction is a constant which is not equal to the grand mean of
the testing set, so it actually **adds** variance.

### Predictions on Evaluation Set
We will use the stepwise-pairwise model to predict `TARGET_WINS` for the
separately provided evaluation set by following the same preprocessing used for
the training and testing sets.

@@

# Code Appendix
The code chunks below represent the code called in order during the analysis.
They are reproduced in the appendix for review and comment.

```{r loadPackages, echo=TRUE, eval=FALSE}
```
```{r readData, echo=TRUE, eval=FALSE}
```
```{r missingTable, echo=TRUE, eval=FALSE}
```
```{r summaryTable, echo=TRUE, eval=FALSE}
```
```{r boxPlot1, echo=TRUE, eval=FALSE}
```
```{r boxPlot2, echo=TRUE, eval=FALSE}
```
```{r corrgram, echo=TRUE, eval=FALSE}
```
```{r testTrain, echo=TRUE, eval=FALSE}
```
```{r imputeMean, echo=TRUE, eval=FALSE}
```
```{r centerScale, echo=TRUE, eval=FALSE}
```
```{r plotScaled, echo=TRUE, eval=FALSE}
```
```{r boxCox, echo=TRUE, eval=FALSE}
```
```{r reScale, echo=TRUE, eval=FALSE}
```
```{r reImputeThenScale, echo=TRUE, eval=FALSE}
```
```{r lm, echo=TRUE, eval=FALSE}
```
```{r nullOutput, echo=TRUE, eval=FALSE}
```
```{r satOutput, echo=TRUE, eval=FALSE}
```
```{r stepOutput, echo=TRUE, eval=FALSE}
```
```{r trainStatTable, echo=TRUE, eval=FALSE}
```
```{r residPlots, echo=TRUE, eval=FALSE}
```
```{r buildTestSet, echo=TRUE, eval=FALSE}
```
```{r predictions, echo=TRUE, eval=FALSE}
```
